\RequirePackage{amsmath}
\documentclass[25pt, a0paper, landscape, margin=15.4mm]{tikzposter}
\def\textmu{}

\usepackage{natbib, textgreek, bussproofs, epigraph, float,
            enumerate, url, graphicx, listings, xfrac}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{filecontents}
\usepackage{color}
\usepackage{fancyvrb}
\usepackage{inconsolata}
\usepackage{tgpagella}
\usepackage{tikz-cd}
\usepackage{adjustbox}
\usepackage{enumitem}
% \sloppy

% \usepackage[usenames,dvipsnames]{xcolor}

\definecolor{IdrisRed}{RGB}{166, 0, 2}
\definecolor{IdrisBlue}{RGB}{0, 0, 182}
\definecolor{IdrisGreen}{RGB}{13, 84, 8}
\definecolor{IdrisPurple}{RGB}{126, 2, 236}

\newcommand{\IdrisData}[1]{\textcolor{IdrisRed}{#1}}
\newcommand{\IdrisType}[1]{\textcolor{IdrisBlue}{#1}}
\newcommand{\IdrisBound}[1]{\textcolor{IdrisPurple}{#1}}
\newcommand{\IdrisFunction}[1]{\textcolor{IdrisGreen}{#1}}
\newcommand{\IdrisMetavar}[1]{\textcolor{cyan}{#1}}
\newcommand{\IdrisKeyword}[1]{{\textbf{#1}}}
\newcommand{\IdrisImplicit}[1]{{\itshape \IdrisBound{#1}}}
\usepackage{setspace, microtype}
\fvset{commandchars=\\\{\},formatcom=\singlespacing, frame=single}
\newcommand{\ty}[1]{\IdrisType{\texttt{#1}}}
\newcommand{\kw}[1]{\IdrisKeyword{\texttt{#1}}}
\newcommand{\fn}[1]{\IdrisFunction{\texttt{#1}}}
\newcommand{\dt}[1]{\IdrisData{\texttt{#1}}}
\newcommand{\bn}[1]{\IdrisBound{\texttt{#1}}}
\newcommand{\cm}[1]{\textcolor{darkgray}{\texttt{#1}}}
\newcommand{\hole}[1]{\IdrisMetavar{\texttt{?}\IdrisMetavar{\texttt{#1}}}}
\newcommand{\Elab}{\ty{Elab}}
\newcommand{\String}{\ty{String}}
\newcommand{\TT}{\ty{TT}}
\newcommand{\Raw}{\ty{Raw}}
\newcommand{\Editorable}{\ty{Editorable}}
\newcommand{\TyDecl}{\ty{TyDecl}}
\newcommand{\FunDefn}{\ty{FunDefn}}
\newcommand{\FunClause}{\ty{FunClause}}

\newcommand{\T}[1]{\texttt{#1}}

\usepackage{more}

\title{Edit-Time Tactics in Idris}
\institute{Wesleyan University} % See Section 4.1
\author{Joomy Korkut}

\usetheme{Simple}  % See Section 5

\begin{document}
% \definecolor{myblue}{RGB}{89,65,242}
\definecolor{myblue}{RGB}{0, 128, 110}
\colorlet{titlebgcolor}{myblue}
\colorlet{innerblocktitlefgcolor}{myblue}
\colorlet{blocktitlefgcolor}{myblue}

\maketitle  % See Section 4.1

\begin{columns}  % See Section 4.4

\column{0.25}

  \block{Abstract\phantom{p}}{
  % \coloredbox[bgcolor=blue,fgcolor=white]{}

  \Large
  \setlength{\parindent}{1.5cm}
  \indent Metaprogramming allows users to write programs that
  write programs. In dependently-typed languages such as \textbf{Idris}, recent work on
  \textbf{elaborator reflection}\cite{elabref} paved the way for new applications of
  metaprogramming by showing that it can be a substitute for \textbf{tactic-based proof
  languages}.  The goal of our work is to use elaborator reflection to write
  \mbox{editor} interaction actions in Idris.

  \indent  Now we can write new tactics as \Elab\ \mbox{actions}
  and run them from the editor, i.e. in \textbf{edit-time}.
  The old editor interaction mode only had a limited number of built-in
  actions, like type-checking holes, case-splitting, and lemma
  extraction. Our work extends its abilities to anything that can be done with
  tactics. We present the design and implementation of this
  feature for the Idris compiler.
}


\column{0.50}
  \block
  {Crash course on the Idris compiler and Idris metaprogramming}{
  \Large
  \scalebox{1.5}{
\newcommand\mlnode[1]{
  \begin{minipage}{8cm}
    \linespread{1}\selectfont \begin{center} \small  #1 \end{center}
  \end{minipage}}
\begin{tikzcd}[ampersand replacement=\&]
  \mlnode{
    Haskell terms\\ \medskip \footnotesize such as the \ty{()} term\\
    \dt{()}
  }
  \arrow[rrr, "\text{\small reflection}" description, bend right=10]
  \&  \&  \&
  \mlnode{
    Haskell representation of Idris core language terms\\ \medskip \scriptsize
    such as the \ty{Term} term\\
    \texttt{\dt{P} (\dt{DCon} \dt{0} \dt{0} \dt{False}) (\dt{UN} \dt{"MkUnit"}) (\dt{P} (\dt{TCon} \dt{0} \dt{0 \dt{False}}) (\dt{UN} \dt{"Unit"}) \dt{Erased})}
  }
  \arrow[lll, "\text{\small reification}" description, bend right=10] \\
  \&  \&  \& \\
  \mlnode{
    Idris terms\\ \medskip \footnotesize such as the \ty{()} term\\
    \dt{()}
  }
  \arrow[r, "\text{\small elaboration}" description, bend right=50]
  \&
  \mlnode{
    Idris core terms\\ \medskip \footnotesize such as the \ty{Unit} term\\
    \dt{MkUnit}
  }
  \arrow[l, "\text{\small delaboration}" description, bend right=50]
  \arrow[rr, "\text{\small quotation}" description, bend right=35]
  \arrow[uurr, "\text{\small \emph{internally represented as}}" description, dashed, bend left=17]
  \& \&
  \mlnode{
    Idris representation of the Haskell representation of Idris core language terms\\
    \medskip \scriptsize such as the \TT\ term\\
    \texttt{\dt{P} (\dt{DCon} \dt{0} \dt{0}) (\dt{UN} \dt{"MkUnit"})\\
    (\dt{P} (\dt{TCon} \dt{0} \dt{0}) (\dt{UN} \dt{"Unit"}) \dt{Erased})}
  }
  \arrow[ll, "\text{\small unquotation}" description, bend right=35]
  \arrow[uu, leftrightarrow, "\text{\small \emph{correspondence}}" description, dashed]
\end{tikzcd}
}
}


\column{0.25}
\block{How it works in Emacs\phantom{p}}{
\Large

  \setlength{\parindent}{1.5cm}
  Suppose we have the following snippet:

  \hspace{-3cm}
  \scalebox{3.5}{
    \begin{minipage}{6cm}
      \VerbatimInput[frame=none]{Before.tex}
    \end{minipage}
  }
  \vspace{1.5cm}

  We want to finish the fill the hole \hole{a} automatically, by
  running the \fn{prover} tactic in the editor.
  Place the cursor on \hole{a} and execute the shortcut we assigned in Emacs
  for \fn{prover}:

  \hspace{-3cm}
  \scalebox{3.5}{
    \begin{minipage}{6cm}
      \VerbatimInput[frame=none]{After.tex}
    \end{minipage}
  }
  \vspace{1.5cm}

  \normalsize
  (\fn{prover} uses Hezarfen, which we describe in applications.)

}


\end{columns}

\begin{columns}
\column{0.25}
\block{The \ty{Editorable} interface\phantom{p}}{
  \Large
  \setlength{\parindent}{1.5cm}
  \indent Editors communicate with the compiler via \mbox{S-expressions}, so
  users should be able \mbox{dictate} how to convey a value of a given type.  We
  define an interface in Idris, which can \mbox{specify} serialization and
  deserialization of the terms of a type into an \ty{SExp} and vice versa.

  % \vspace{-0.2cm}
  \hspace{-3cm}
  \scalebox{3.5}{
    \begin{minipage}{6cm}
      \VerbatimInput[frame=none]{Editorable.tex}
    \end{minipage}
  }
  \vspace{1.5cm}

  \indent Usually we can define \ty{Editorable} implementations in Idris
  itself. But for Idris types that represent the Haskell representation of
  Idris core terms, such as \ty{TT} and \ty{TyDecl}, we need to define
  \ty{Editorable} implementations primitively in order to be able to parse,
  \mbox{elaborate}, delaborate, and pretty print them.
}

\column{0.25}
\block{Type-checking with \ty{Editorable}\phantom{p}}{
\Large
  \setlength{\parindent}{1.5cm}

  \indent We want to ascertain during type-checking that an \Elab\ action we
  will use as an \mbox{editor} action is suitable for that task. That \mbox{requires}
  checking that all components of the type signature have an implementation of
  \ty{Editorable}.  Let's introduce a new function modifier \mbox{syntax}
  \kw{\%editor}, which enforces this constraint in compile-time.

  % \vspace{\baselineskip}
  \vspace{0.35cm}

  \indent If we have a function\\
  \indent \texttt{\fn{toy} : \ty{TTName} -> \ty{TT} -> \ty{Elab} \ty{TT}}\\
  and we declare it an editor action with \kw{\%editor}, then the types
  \ty{TTName}, \ty{TT} (from the argument), and \ty{TT} (extracted from the
  return type) must have implementations of \ty{Editorable}, so that we are
  sure we know how to deserialize the arguments to \fn{toy} and \mbox{serialize} its result.
}

\column{0.25}
\block{Toy example}{
\Large
  \setlength{\parindent}{1.5cm}
  \vspace{-1.7cm}
  \hspace{-3cm}
  \scalebox{3.5}{
    \begin{minipage}{6cm}
      \VerbatimInput[frame=none]{Toy.tex}
    \end{minipage}
  }
  \vspace{1.5cm}

  \indent We can run \fn{toy} from the editor with a hole name and
  some string representing a \TT\ term. If the type of the hole is \ty{Nat},
  then \fn{toy} returns the term it was given, otherwise fails.

  \indent Idris code cannot tell Emacs to replace the name under cursor with
  some expression; we have to write 5-10 lines of trivial Emacs Lisp code
  to glue the \Elab\ action to our editor.
  \mbox{\emph{(If we are using Emacs, of course.)}}
}

\column{0.25}
\block{Useful applications}{
\Large
  \vspace{-0.7cm}
  \begin{enumerate}[labelsep=0.5cm]
    \renewcommand{\labelenumi}{(\arabic{enumi})}
    \item Replace the existing built-in editor actions (written in Haskell as a
      part of the compiler) with edit-time tactics (written in Idris). In our
      work, we present a replacement for the ``add clause'' action.
    \item Write a proof search mechanism better than the built-in one, as an
      edit-time tactic. In our work, we present \textbf{Hezarfen}, a theorem
      prover tactic that decides intuitionistic propositional logic, generates
      \TT\ proof terms, and simplifies the proof terms as much as possible.
\end{enumerate}
}

\block[titleoffsety=3cm, bodyoffsety=3cm]
  {References}{
  \vspace{-0.5cm}
  \bibliographystyle{plain}
  \renewcommand{\section}[2]{}%
  \bibliography{../thesis/paper}
}
\end{columns}


\end{document}
