\documentclass[sigplan, review, timestamp]{acmart}

% Copyright
%\setcopyright{none}
%\setcopyright{acmcopyright}
\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\setcopyright{usgov}
%\setcopyright{usgovmixed}
%\setcopyright{cagov}
%\setcopyright{cagovmixed}

% DOI
\acmDOI{}

% ISBN
\acmISBN{}

%Conference
\acmConference[TyDe'18]{the 3rd ACM SIGPLAN International Workshop on Type-Driven Development}{September 27, 2018}{St. Louis, MO, USA}
\acmYear{2018}
\copyrightyear{2018}
\acmPrice{15.00}
\acmSubmissionID{}

\usepackage{enumerate, enumitem}
\usepackage{fancyvrb}
\fvset{commandchars=\\\{\}}
\usepackage{xcolor}
\usepackage{setspace}
\usepackage{todonotes}
\usepackage{tikz-cd}
\usepackage{adjustbox}
% Idris
% in REPL run `:pp latex 80 <expr>` to get the colored verbatim

\usepackage{amssymb}
\newcommand\fakeslant[1]{%
  \special{pdf: literal 1 0 0.167 1 0 0 cm}#1\special{pdf: literal 1 0 -0.167 1 0 0 cm}}
\definecolor{IdrisRed}{HTML}{BC0045}
\definecolor{IdrisBlue}{HTML}{0000d6}
\definecolor{IdrisGreen}{HTML}{008b00}
\definecolor{IdrisLilac}{HTML}{AC22BF}
\definecolor{IdrisYellow}{HTML}{E16800}
\definecolor{IdrisLightBlue}{HTML}{A4CDFF}
\newcommand{\IdrisData}[1]{\textcolor{IdrisRed}{#1}}
\newcommand{\IdrisType}[1]{\textcolor{IdrisBlue}{#1}}
\newcommand{\IdrisBound}[1]{\textcolor{IdrisLilac}{\fakeslant{#1}}}
\newcommand{\IdrisFunction}[1]{\textcolor{IdrisGreen}{#1}}
\newcommand{\IdrisMetavar}[1]{\textcolor{IdrisYellow}{#1}}
\newcommand{\IdrisKeyword}[1]{{\textbf{#1}}}
\newcommand{\IdrisImplicit}[1]{{\itshape \IdrisBound{#1}}}
\newcommand{\ty}[1]{\IdrisType{\texttt{#1}}}
\newcommand{\kw}[1]{\IdrisKeyword{\texttt{#1}}}
\newcommand{\fn}[1]{\IdrisFunction{\texttt{#1}}}
\newcommand{\dt}[1]{\IdrisData{\texttt{#1}}}
\newcommand{\bn}[1]{\IdrisBound{\texttt{#1}}}
\newcommand{\cm}[1]{\textcolor{darkgray}{\texttt{#1}}}
\newcommand{\hole}[1]{\IdrisMetavar{\texttt{?}\IdrisMetavar{\texttt{#1}}}}
\newcommand{\Elab}{\ty{Elab}}
\newcommand{\IO}{\ty{IO}}
\newcommand{\String}{\ty{String}}
\newcommand{\TT}{\ty{TT}}
\newcommand{\Raw}{\ty{Raw}}
\newcommand{\Editorable}{\ty{Editorable}}
\newcommand{\TyDecl}{\ty{TyDecl}}
\newcommand{\FunDefn}{\ty{FunDefn}}
\newcommand{\FunClause}{\ty{FunClause}}
\newcommand{\sexp}{\mbox{S-expression}} %don't hyphenate
\newcommand{\mt}[1]{\mbox{\texttt{#1}}}
\newcommand{\qt}[1]{\kw{\`{}(}{#1}\kw{)}} %quotation `(...)
\newcommand{\antiqt}{\kw{\textasciitilde}} %antiquotation ~

% Replicating selecting regions in the editor:
% \newcommand{\select}[1]{\fcolorbox{black}{IdrisLightBlue}{#1}}
% A version of select without any space around the box.
% Hard to see in black & white so I chose the one with the space
\newcommand{\select}[1]{{\setlength{\fboxsep}{0.05cm}\fcolorbox{black}{IdrisLightBlue}{#1}}}

\newcommand{\Red}[1]{{\color{red} #1}}
\newcommand{\TODO}[1]{{\color{red}{[TODO: #1]}}}
\newcommand{\FYI}[1]{{\color{green}{[FYI: #1]}}}
\newcommand{\forceindent}{\hspace{\parindent}}

\usepackage[normalem]{ulem}
\makeatletter
\newcommand\colorwave[1][blue]{\bgroup \markoverwith{\lower3.5\p@\hbox{\sixly \textcolor{#1}{\char58}}}\ULon}
\font\sixly=lasy6 % does not re-load if already loaded, so no memory problem.
\makeatother
\newcommand{\IdrisError}[1]{\colorwave[red]{#1}}

\newcommand\mlnode[1]{
  \begin{minipage}{3.3cm}
    \linespread{1}\selectfont \begin{center}\small #1 \end{center}
  \end{minipage}}

\sloppy

%% Get rid of sloppy line breaks in \citet citations
\bibpunct{\nolinebreak[4][}{]}{,}{n}{}{,}
\makeatletter
\renewcommand*{\NAT@spacechar}{~}
\makeatother

\begin{document}
\title[Extensible Type-Directed Editing]{Extensible Type-Directed Editing}


\author{Joomy Korkut}
\orcid{0000-0001-6784-7108}
%\authornote{Author's current affiliation: Princeton University}
\affiliation{
  \institution{Wesleyan University}
  % \city{Middletown}
  % \state{Connecticut}
  % \country{USA}
}
\email{ckorkut@wesleyan.edu}

\author{David Thrane Christiansen}
\affiliation{
  \institution{Galois, Inc.}
  % \city{Portland}
  % \state{Oregon}
  % \country{USA}
}
\email{dtc@galois.com}


% The default list of authors is too long for headers.
\renewcommand{\shortauthors}{Joomy Korkut and David Christiansen}

\begin{abstract}
  Dependently typed programming languages, such as Idris and Agda, feature rich interactive environments that use informative types to assist users with the construction of programs. However, these interactive environments have been provided by the authors of the language, and users have not had an easy way to extend and customize them. We address this problem by extending Idris's metaprogramming facilities with primitives for describing new type-directed editing features, making Idris's editors as extensible as its elaborator.
\end{abstract}

%
% The code below should be generated by the tool at
% http://dl.acm.org/ccs.cfm
% Please copy and paste the code instead of the example below.
%
 \begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10011007.10011006.10011008.10011009.10011012</concept_id>
<concept_desc>Software and its engineering~Functional languages</concept_desc>
<concept_significance>300</concept_significance>
</concept>
<concept>
<concept_id>10011007.10011006.10011008.10011024</concept_id>
<concept_desc>Software and its engineering~Language features</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}
\ccsdesc[300]{Software and its engineering~Functional languages}
\ccsdesc[300]{Software and its engineering~Language features}

\keywords{Metaprogramming, dependent types, editors}

\maketitle

\emph{This paper uses colors in the example code.}
\input{introduction}
\input{design}
\input{implementation}
\input{applications}
\input{relatedwork}
\input{conclusion}

\section*{Acknowledgments}

We would like to thank Daniel R.\ Licata for his extensive support
while advising the first author's M.A.\
thesis~\citep{editTimeTactics}, on which this paper is largely based.
The first author was generously supported by Wesleyan University.  The
second author began working on this project during a postdoctoral fellowship at
Indiana University, where he was supported by the United States
National Science Foundation grant 1540276. Galois, Inc.\ has
generously supported further work.
We would also like to thank Daniel P.\ Friedman and Austin Tasato for their
comments on our draft.


\bibliographystyle{ACM-Reference-Format}
\bibliography{../thesis/paper}

\end{document}
