\documentclass[11pt, ma]{westhesis}
\setcounter{tocdepth}{4}
\setcounter{secnumdepth}{4}
\usepackage[utf8]{inputenc}
\usepackage{amsmath, amsfonts, amssymb, bussproofs, listings, float, upgreek, stmaryrd, epigraph, afterpage}
\usepackage[square,sort&compress,numbers]{natbib}
\usepackage[font=small]{caption}
\usepackage{fancyvrb}
\usepackage[usenames,dvipsnames]{xcolor}
% \usepackage[usenames,dvipsnames,monochrome]{xcolor} %black and white
\usepackage[T1]{tipa}
\usepackage{graphicx}
\graphicspath{ {images/} }
\usepackage{tgpagella}
\usepackage{inconsolata}
\usepackage{ragged2e}
\usepackage[hang, flushmargin]{footmisc}
\usepackage[backref=page,hyperfootnotes=false]{hyperref}
\usepackage{footnotebackref}
\definecolor{IdrisRed}{RGB}{166, 0, 2}
\definecolor{IdrisBlue}{RGB}{0, 0, 182}
\definecolor{IdrisGreen}{RGB}{13, 84, 8}
\definecolor{IdrisPurple}{RGB}{126, 2, 236}
\hypersetup{
  colorlinks = true,
  linkcolor = {blue},
  citecolor = {blue},
  urlcolor = {blue},
  anchorcolor = {blue}
}
\usepackage{tikz-cd}
\usepackage{adjustbox}
\usepackage{enumerate}

% Idris
% in REPL run `:pp latex 80 <expr>` to get the colored verbatim
\newcommand{\IdrisData}[1]{\textcolor{IdrisRed}{#1}}
\newcommand{\IdrisType}[1]{\textcolor{IdrisBlue}{#1}}
\newcommand{\IdrisBound}[1]{\textcolor{IdrisPurple}{#1}}
\newcommand{\IdrisFunction}[1]{\textcolor{IdrisGreen}{#1}}
\newcommand{\IdrisMetavar}[1]{\textcolor{cyan}{#1}}
\newcommand{\IdrisKeyword}[1]{{\textbf{#1}}}
\newcommand{\IdrisImplicit}[1]{{\itshape \IdrisBound{#1}}}
\usepackage{setspace, microtype}
\fvset{commandchars=\\\{\},formatcom=\singlespacing, frame=single}
\newcommand{\ty}[1]{\IdrisType{\texttt{#1}}}
\newcommand{\kw}[1]{\IdrisKeyword{\texttt{#1}}}
\newcommand{\fn}[1]{\IdrisFunction{\texttt{#1}}}
\newcommand{\dt}[1]{\IdrisData{\texttt{#1}}}
\newcommand{\bn}[1]{\IdrisBound{\texttt{#1}}}
\newcommand{\cm}[1]{\textcolor{darkgray}{\texttt{#1}}}
\newcommand{\hole}[1]{\IdrisMetavar{\texttt{?}\IdrisMetavar{\texttt{#1}}}}
\newcommand{\Elab}{\ty{Elab}}
\newcommand{\String}{\ty{String}}
\newcommand{\TT}{\ty{TT}}
\newcommand{\Raw}{\ty{Raw}}
\newcommand{\Editorable}{\ty{Editorable}}
\newcommand{\TyDecl}{\ty{TyDecl}}
\newcommand{\FunDefn}{\ty{FunDefn}}
\newcommand{\FunClause}{\ty{FunClause}}

\newcommand{\quadthree}{\qquad\quad}
\newcommand{\quadfour}{\quadthree\quad}
\newcommand{\quadfive}{\quadfour\quad}
\newcommand{\quadsix}{\quadfive\quad}
\newcommand{\quadseven}{\quadsix\quad}
\newcommand{\quadeight}{\quadseven\quad}
\newcommand{\quadten}{\quadfive\quadfive}
\newcommand{\Red}[1]{{\color{red} #1}}
\newcommand{\TODO}[1]{{\color{red}{[TODO: #1]}}}
\newcommand{\FYI}[1]{{\color{green}{[FYI: #1]}}}
\newcommand{\forceindent}{\hspace{\parindent}}
\sloppy


% \allowdisplaybreaks

\department{Mathematics and Computer Science}
\submitdate{May 2018}
\advisor{Daniel R. Licata}
\title{Edit-Time Tactics in Idris}
\author{Joomy Korkut}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[chapter]
\newtheorem{corollary}{Corollary}[theorem]
\newtheorem{lemma}[theorem]{Lemma}

\theoremstyle{definition}
\newtheorem{defn}{Definition}[chapter]

\newcommand{\T}[1]{\texttt{#1}}
\lstset{mathescape=true,basicstyle=\ttfamily}

\newcommand\numberthis{\addtocounter{equation}{1}\tag{\theequation}}

\numberwithin{section}{chapter}
\numberwithin{figure}{chapter}

\begin{document}

\begin{abstract}
  \forceindent Metaprogramming is a technique that allows users to write
  programs that write programs. In dependently-typed languages such as Idris,
  recent work on elaborator reflection\cite{elabref} paved the way for new
  applications of metaprogramming by showing that it can be a substitute for
  tactic-based proof languages. The goal of our work is to use elaborator
  reflection to write editor interaction actions in Idris.

  Currently in Idris modes of editors such as Emacs, users can perform actions
  like type-checking holes, case-splitting, and lemma extraction.
  Implementations of all of these Idris editor actions are hard-coded in the
  compiler and they are written in Haskell. Our work will allow us to rewrite
  them in Idris as metaprograms, and to move them into an Idris library,
  instead of having them embedded into the compiler.

  Furthermore, we can write our own tactics through elaborator
  reflection, and run them from the editor, i.e. in edit-time.
  This would extend the abilities of the editor interaction mode from the
  current built-in features to anything that can be done with tactics.
  In our work, we present the design and implementation of this feature in the
  Idris compiler.

  We also implement an intuitionistic theorem prover tactic that is meant to be an
  better alternative to the built-in proof search editor action, and an
  add-clause tactic that exemplifies how we can move some of the hard-coded
  features from the compiler to a library.
\end{abstract}

\begin{dedication}
  \epigraph{``What do compilers do? They manipulate programs! Making it easy
  for users to manipulate their own programs, and also easy to interlace their
  manipulations with the compiler’s manipulations, creates a powerful new
  tool.''}{\textit{Tim Sheard and Simon Peyton Jones}\cite{th}}
\end{dedication}

\begin{acknowledgements}
\forceindent First and foremost, I would like to thank Dan Licata, my research
  advisor.  Even though the topic I picked was not very close to the work he
  does, he played along and gave me very helpful advice whenever I got stuck.
  Since this project also had a user interface aspect, his insight as a
  long-time user of proof assistants shaped my decisions.  He has had more
  impact on my research taste than anyone, and I am indebted to him for his
  kindness and patience.

I am incredibly fortunate to have met David Christiansen last summer and talked
to him about his work on metaprogramming for dependent types.
In his dissertation\cite{davidphd} he laid the foundation that makes this
project possible, and he later pitched me the idea of using elaborator
reflection for interactive editing.
I could not have finished this work without him putting up with my incessant
stream of questions. (My words, not his.)

I would like to express my gratitude to James Lipton and Edward Morehouse for
reading and evaluating my thesis. I would also like to thank Cyrus Omar for
coining the term ``edit-time tactics''. Even though he did so in a different
context\cite{hazelnut}, I cannot think of a better way to describe this
project. I would like to thank Matthew Wilson for reading my draft many
times, and helping me put my project in perspective with our conversations
about the history and future of IDEs and programming.

I want to thank Joseph Cutler and Mitchell Riley for the programming languages
discussions we had over the year; I am grateful for our little PL community.

Finally, I am thankful to Yulia, Mehmet, Recep, Kivanc, Isin Ekin and
my family for their emotional support. This has not been an easy year for me,
but I made it through thanks to all of you.
\end{acknowledgements}

\frontmatter
\maketitle
\makeabstract
\makededication
\makeack

\tableofcontents

\newcommand{\nocontentsline}[3]{}
\bgroup\let\addcontentsline=\nocontentsline
\renewcommand\numberline[1]{#1.\ }
\listoffigures
\egroup

\mainmatter

% \begin{figure}[ht]
% \caption{Some Idris code}
% \label{code:idrisExample}
% \begin{Verbatim}

% \end{Verbatim}
% \end{figure}

\include{introduction}
\include{background}
\include{design}
\include{implementation}
\include{applications}
\include{relatedwork}
\include{conclusion}
\bibliography{paper}
\bibliographystyle{plain}
\end{document}
